---
title: "Homework 1"
author: "Christopher Bussen"
date: "2023-01-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Character - 1

A character object is used to store text, letters, or words (strings) in R.

```{r}
x <- "Hello"
y <- "UD students!"
class(x) # we can use class() function to obtain the data type
```

```{r}
nchar(x) # use nchar() to count the number of characters
```

**Note:**When defining strings, double quotes " " and single quotes ' ' are interchangeably but double quotes are preferred (and character constants are printed using double quotes), so single quotes are normally only used to delimit character constants containing double quotes R Documentation ([2020](https://ying-ju.github.io/teaching/MTH209/Lesson2.html#ref-rquotes2020)).

## Character - 2
If we want to combine two strings into one string, we can use paste() or paste0() function.

```{r}
paste(x,y)
```

```{r}
paste(x,y,sep=",")
```

```{r}
paste(x,y,sep=", ")
```

```{r}
paste(x, ", ", y)
```

```{r}
paste0(x,y)
```

These two functions could be very useful.

## Character - 3
Here we give one advanced example.

```{r}
allfiles1 <- paste("file_", 1:5)
allfiles2 <- paste("file_", 1:5, collapse = "_")
allfiles3 <- paste("file", 1:5, sep = "_")
allfiles1
```

```{r}
allfiles2
```

```{r}
allfiles3
```

## Integer
An integer object is used to store numeric data without decimals.

```{r}
x2 <- c(-3.13, 2.47, 6, -1.5, 4.29, 2.72, 1, 0, 3.85)
x3 <- as.integer(x2) # only remain the integers
x3
```

```{r}
class(x3)
```

